---
title: "Module 2: Chi-Square testing and ANOVA"
author: "Sankalp Biswal, ALY6015 Intermediate Analytics"
date: "2024-01-20"
output: html_document
---

# Introduction

This report explores the application of Chi-Square testing and ANOVA to investigate various real-world datasets. The module focuses on assessing relationships, differences, and dependencies across multiple domains, including medical research, airline performance, movie industry trends, military personnel statistics, food industry analysis, and agricultural studies. The statistical methods employed provide insights into how variables interact and differ across groups, offering valuable conclusions from the data.

------------------------------------------------------------------------

\* Did not calculate critical value for Task 1 - 7 since decision was made by comparing p-value and significance level.

------------------------------------------------------------------------

# Task 1(Section 11-16. Blood types)

\
A medical researcher wishes to see if hospital patients in a large hospital have the same blood type\
distribution as those in the general population. The distribution for the general population is as follows:\
type A, 20%; type B, 28%; type O, 36%; and type AB = 16%. He selects a random sample of 50 patients and\
finds the following: 12 have type A blood, 8 have type B, 24 have type O, and 6 have type AB blood.\
At α = 0.10, can it be concluded that the distribution is the same as that of the general population?

# Solution

## $\chi^2$ Test for Goodness of Fit

Form the vectors of $expected$ probabilities and $observed$ frequencies, then run the test.

```{r}
sample_size <- 50
expected    <- c(0.20, 0.28, 0.36, 0.16)
observed    <- c(12, 8, 24, 6)
```

### Step 1 -- State the hypothesis and identify the claim

-   **Null hypothesis:** The blood type distribution in hospital patients is the same as that in the general population.

    **H0**: ${ 𝑃_A=0.20, 𝑃_B=0.28, 𝑃_O=0.36, 𝑃_AB=0.16}$

-   **Alternative hypothesis:** The blood type distribution in hospital patients is different from that in the general population.

    **H1**: ${ 𝑃_A\neq0.20 \text{ or } 𝑃_B\neq0.28 \text{ or } 𝑃_O\neq0.36 \text{ or } 𝑃_AB\neq0.16}$ i.e. the distribution is not the same as stated in the null hypothesis.

### Step 2 - Run the test all the way

```{r}
chisq.test(observed, p = expected, correct = FALSE) # not to apply continuity correction
```

### Step 3 - Make the decision to reject or not reject the null hypothesis

The decision rule is that if the p-value is greater than α, we do not have sufficient evidence to reject the null hypothesis.

$p$-value \> $\alpha$ i.e. 0.1404 $>$ 0.05,so we don't reject **H0**

### Step 4 - Summary

The evidence leads us to the conclusion that there is not enough evidence to reject the null hypothesis. So, the blood type distributions do not differ from the general population.

------------------------------------------------------------------------

# Task 2 (Section 11-1. 8. On-Time Performance by Airlines)

\
According to the Bureau of Transportation Statistics, on-time performance by the airlines is described\
as follows:

| Action                                          | \% of Time |
|-------------------------------------------------|------------|
| On-time                                         | 70.8       |
| National Aviation System delay                  | 8.2        |
| Aircraft arriving late                          | 9.0        |
| Other (because of weather and other conditions) | 12.0       |

\
Records of 200 randomly selected flights for a major airline company showed that 125 planes were on time; 40 were delayed because of weather, 10 because of a National Aviation System delay, and the rest because of arriving late. At α = 0.05, do these results differ from the government's statistics?\

Source: www.transtats.bts.gov

# Solution

## $\chi^2$ Test for Goodness of Fit

Form the vectors of $expected$ probabilities and $observed$ frequencies, then run the test.

```{r}
observed <- c(125, 40, 10, 25)  # 25 is for "Aircraft arriving late"

expected <- c(0.708, 0.082, 0.09, 0.12)
```

### Step 1 - State the hypothesis and identify the claim

-   **Null Hypothesis (H0):** The distribution of flight delay reasons for the airline matches the government's statistics.
-   **Alternative Hypothesis (H1):** The distribution of flight delay reasons for the airline differs from the government's statistics.

### Step 2 - Run the test all the way

```{r}
#In order to convert scientific notation
options(scipen = 999)

chisq.test(observed, p = expected, correct = FALSE) # not to apply continuity correction
```

### Step 3 - Make the decision to reject or not reject the null hypothesis

The decision rule is that when the $p$-value is less than $\alpha$ we should reject the null hypothesis.

$p$-value \< $\alpha$ i.e. 0.00000001357 $<$ 0.05, so we should reject **H0**

### Step 4 - Summary

The evidence leads us to the conclusion that we should reject the null hypothesis at 5% significance level. So, the flight time distributions differ from the government statistics.

------------------------------------------------------------------------

# Task 3 (Section 11-2.8. Ethnicity and Movie Admissions)

Are movie admissions related to ethnicity? A 2014 study indicated the following numbers of admissions (in thousands) for two different years. At the 0.05 significance level, can it be concluded that movie attendance by year depended on ethnicity?

|      | Caucasion | Hispanic | African American | Other |
|------|-----------|----------|------------------|-------|
| 2013 | 724       | 335      | 174              | 107   |
| 2014 | 370       | 292      | 152              | 140   |

Source: MPAA Study

# Solution

## $\chi^2$ Test of Independence

Form matrix from the above table

```{r}
df_3 <- matrix(c(724, 335,174,107,370,292,152,140), 
                   nrow = 2, ncol = 4, byrow = TRUE,
                   dimnames = list(c("2013", "2014"), 
                                   c("Caucasian", "Hispanic", "African American", "Other")))

```

### Step 1 - State the hypothesis and identify the claim

-   **Null Hypothesis (H0):** There is no association between movie attendance and ethnicity. This means that movie attendance is independent of ethnicity across the different years.
-   **Alternative Hypothesis (H1):** There is an association between movie attendance and ethnicity. This implies that movie attendance depends on ethnicity, and the patterns of attendance vary between different ethnic groups across the years.

### Step 2 - Run the test all the way

```{r}
chisq.test(df_3)
```

### Step 3 - Make the decision to reject or not reject the null hypothesis

The decision rule is that when the $p$-value is less than $\alpha$ we should reject the null hypothesis.

$p$-value \< $\alpha$ i.e. 0.0000000000005478 $<$ 0.05,so we should reject $\it{H_{0}}$

### Step 4 - Summary

The evidence leads us to the conclusion that we should reject the null hypothesis at 5% significance level. So, the movie attendance depends on ethnicity.

------------------------------------------------------------------------

# Task 4 (Section 11-2.10 Women in the Military)

This table lists the numbers of officers and enlisted personnel for women in the military. At α = 0.05, is there sufficient evidence to conclude that a relationship exists between rank and branch of the Armed Forces?

|              | Officers | Enlisted |
|--------------|----------|----------|
| Army         | 10,791   | 62,491   |
| Navy         | 7,816    | 42,750   |
| Marine Corps | 932      | 9,525    |
| Air Force    | 11,819   | 54,344   |

Source: New York Times Almanac

# Solution

## $\chi^2$ Test for Goodness of Fit

```{r}
df_4 <- matrix(c(10791, 62491,   # Army
                          7816, 42750,    # Navy
                          932, 9525,      # Marine Corps
                          11819, 54344),  # Air Force
                        nrow = 4, ncol = 2, byrow = TRUE,
                        dimnames = list(c("Army", "Navy", "Marine Corps", "Air Force"),
                                        c("Officers", "Enlisted")))

```

### Step 1 - State the hypothesis and identify the claim

-   **Null Hypothesis (H0):** There is no relationship between rank and branch of the Armed Forces. This means the proportion of officers and enlisted personnel is the same across different branches.
-   **Alternative Hypothesis (H1):** There is a relationship between rank and branch of the Armed Forces. This means the proportion of officers and enlisted personnel varies across different branches.

### Step 2 - Run the test all the way

```{r}
chisq.test(df_4)
```

### Step 3 - Make the decision to reject or not reject the null hypothesis

The decision rule is that when the $p$-value is less than $\alpha$ we should reject the null hypothesis.

$p$-value \< $\alpha$ i.e. 0.00000000000000022 $<$ 0.05,so we should reject $\it{H_{0}}$

### Step 4 - Summary

The evidence leads us to the conclusion to reject the null hypothesis. So, there is sufficient evidence to conclude that there is relation between rank and branch of the armed forces.

------------------------------------------------------------------------

# Task 5 (Section 12-1.8 Sodium Contents of Foods)

Perform a complete one-way ANOVA. If the null hypothesis is rejected, use either the Scheffé or Tukey test to see if there is a significant difference in the pairs of means. Assume all assumptions are met.

The amount of sodium (in milligrams) in one serving for a random sample of three different kinds of foods is listed. At the 0.05 significance level, is there sufficient evidence to conclude that a difference in mean sodium amounts exists among condiments, cereals, and desserts?

## Dataset

| Condiments | Cereals | Desserts |
|------------|---------|----------|
| 270        | 260     | 100      |
| 130        | 220     | 180      |
| 230        | 290     | 250      |
| 180        | 290     | 250      |
| 80         | 200     | 300      |
| 70         | 320     | 360      |
| 200        | 140     | 300      |
|            |         | 160      |

Source: The Doctor's Pocket Calorie, Fat, and Carbohydrate Counter

# Solution

## ANOVA (One Way) Test

```{r}
alpha         <- 0.05
condiments    <- c(270,130,230,180,80,70,200)
cereals         <- c(260,220,290,290,200,320,140)
desserts        <- c(100,180,250,250,300,360,300,160)
```

Form the dataframe (table) from the row vectors

```{r}
food <- matrix(c(rep("Condiments", 7), rep("Cereals", 7), rep("Desserts", 8), condiments , 
                  cereals, desserts),ncol=2)
df_5 <- data.frame(food)
names(df_5) <- c('food_type', 'sodium_milligrams')
df_5$sodium_milligrams <- as.numeric(as.character(df_5$sodium_milligrams)) # need to convert "sodium(in milligrams)" from factor to numeric
df_5
```

### Step 1 - State the hypothesis

-   **Null hypothesis:** There is no significant difference in the mean sodium contents among condiments, cereals, and desserts.

    **(H0)**: $\mu_1=\mu_2=\mu_3$

-   **Alternative hypothesis:** (**H1)**: At least one group mean is significantly different from the others. This means there is a significant difference in the mean sodium contents among at least one pair of the groups (condiments, cereals, and desserts).

### Step 2 - Run the test all the way

Using `aov()` as shown below.

```{r}
anova_one_way <- aov(as.numeric(sodium_milligrams)~food_type, data=df_5)
summary(anova_one_way)
```

### Step 3 - Make the decision to reject or not reject the null hypothesis

The $p$-value is greater than the usual threshold of $\alpha$=0.05. We are confident to say there is no a statistical difference between the groups. Therefore we fail to reject the null hypothesis.

$p$-value of 0.118 \> 0.05,so we fail to reject the Null hypothesis.

### Step 4 - Summary

At a significance level of 0.05, there is not enough evidence to conclude that significant differences exist in the mean sodium amounts among the different food types. Therefore, we fail to reject the null hypothesis that the means are equal across food types.

------------------------------------------------------------------------

# Task 6 (Section 12-2.10 Sales for Leading Companies)

Perform a complete one-way ANOVA. If the null hypothesis is rejected, use either the Scheffé or Tukey test to see if there is a significant difference in the pairs of means. Assume all assumptions are met.

The sales in millions of dollars for a year of a sample of leading companies are shown. At α = 0.01, is there a significant difference in the means?

## Dataset

+--------+-----------+--------+
| Cereal | Chocolate | Coffee |
|        |           |        |
|        | Candy     |        |
+--------+-----------+--------+
| 578    | 311       | 261    |
+--------+-----------+--------+
| 320    | 106       | 185    |
+--------+-----------+--------+
| 264    | 109       | 302    |
+--------+-----------+--------+
| 249    | 125       | 689    |
+--------+-----------+--------+
| 237    | 173       |        |
+--------+-----------+--------+

: Source: Information Resources, Inc.

# Solution

## ANOVA (One Way) Test

```{r}
alpha         <- 0.01
cereal    <- c(578, 320, 264, 249,237)
candy         <- c(311, 106, 109, 125,173)
coffee        <- c(261, 185, 302,689)
```

Form the dataframe (table) from the row vectors

```{r}
Task6 <- matrix(c(rep("Cereal", 5), rep("Candy", 5), rep("Coffee", 4), cereal , 
                  candy, coffee),ncol=2)
df_6 <- data.frame(Task6)
names(df_6) <- c('sample_type', 'sales_million_dollars')
df_6$sales_million_dollars <- as.numeric(as.character(df_6$sales_million_dollars)) # need to convert "sales_million_dollars" from factor to numeric
df_6
```

### Step 1 - State the hypothesis

-   **Null hypothesis :** There is no significant difference in the mean sales among the four categories (Chocolate, Cereal, Candy, Coffee).

    **H0**: $\mu_1=\mu_2=\mu_3$

-   **Alternative hypothesis :** There is a significant difference in the mean sales among at least one pair of the four categories.

    **H1**: At least one mean is different from the others

### Step 2 - Run the test all the way

Using `aov()` as shown below.

```{r}
anova_one_way <- aov(as.numeric(sales_million_dollars)~sample_type, data=df_6)
summary(anova_one_way)
```

### Step 3 - Make the decision to reject or not reject the null hypothesis

The $p$-value is greater than the desired threshold of $\alpha$=0.01. We are confident to say there is no a statistical difference between the groups. Therefore we fail to reject the null hypothesis.

$p$-value of 0.16 \> 0.01,so we fail to reject the Null hypothesis.

### Step 4 - Summary

At a significance level of 0.01, there is not enough evidence to conclude that significant differences exist in the mean "sales in million dollars" among the different sample types. Therefore, we fail to reject the null hypothesis that the means are equal across sample types.

------------------------------------------------------------------------

# Task 7 (Section 12-2.12 Per-Pupil Expenditures)

Perform a complete one-way ANOVA. If the null hypothesis is rejected, use either the Scheffé or Tukey test to see if there is a significant difference in the pairs of means. Assume all assumptions are met.

The expenditures (in dollars) per pupil for states in three sections of the country are listed. Using α =0.05, can you conclude that there is a difference in means?

## Dataset

| Eastern Third | Middle Third | Western Third |
|---------------|--------------|---------------|
| 4946          | 6149         | 5282          |
| 5953          | 7451         | 8605          |
| 6202          | 6000         | 6528          |
| 7243          | 6479         | 6911          |
| 6113          |              |               |

: Source: New York Times Almanac

```{r}
alpha         <- 0.05
eastern    <- c(4946,5953,6202,7243,6113)
middle         <- c(6149,7451,6000,6479)
western        <- c(5282,8605,6528,6911)
```

# Solution

## ANOVA (One Way) Test

Form the dataframe (table) from the row vectors

```{r}
task7 <- matrix(c(rep("eastern", 5), rep("middle", 4), rep("western", 4), eastern , 
                  middle, western),ncol=2)
df_7 <- data.frame(task7)
names(df_7) <- c('country_section', 'expenditure_dollars_per_pupil')
df_7$expenditure_dollars_per_pupil <- as.numeric(as.character(df_7$expenditure_dollars_per_pupil)) # need to convert "mpg" from factor to numeric
df_7
```

### Step 1 - State the hypothesis.

-   **Null hypothesis:** There is no significant difference in the mean expenditures per pupil among the Eastern Third, Middle Third, and Western Third sections of the country.

    **H0**: $\mu_1=\mu_2=\mu_3$

-   **Alternative hypothesis:** There is a significant difference in the mean expenditures per pupil among at least one pair of the three sections.

    **H1**: At least one mean is different from the others

### Step 2 - Run the test all the way

Using `aov()` as shown below.

```{r}
anova_one_way <- aov(as.numeric(expenditure_dollars_per_pupil)~country_section, data=df_7)
summary(anova_one_way)
```

### Step 3 - Make the decision to reject or not reject the null hypothesis

The $p$-value is greater than the usual threshold of $\alpha$=0.05. We are confident to say there is no a statistical difference between the groups. Therefore we fail to reject the null hypothesis.

$p$-value of 0.543 \> 0.05,so we fail to reject Null hypothesis

### Step 4 - Summary

At a significance level of 0.05, there is not enough evidence to conclude that significant differences exist in the mean "expenditure in dollars" among the different country sections. Therefore, we fail to reject the null hypothesis that the means are equal across the different country sections.

------------------------------------------------------------------------

# Task 8 (Section 12-3.10 Increasing Plant Growth)

A gardening company is testing new ways to improve plant growth. Twelve plants are randomly selected and exposed to a combination of two factors:

• a "Grow-light" in two different strengths.

• a plant food supplement with different mineral supplements.

After several days, the plants are measured for growth, and the results (in inches) are put into the appropriate boxes.

|              | Grow Light 1  | Grow Light 2  |
|--------------|---------------|---------------|
| Plant Food A | 9.2, 9.4, 8.9 | 8.5, 9.2, 8.9 |
| Plant Food B | 7.1, 7.2, 8.5 | 5.5, 5.8, 7.6 |

# Solution

## Two-way ANOVA Test

```{r}

# Growth data for plants
grow_light_1_plant_food_A <- c(9.2, 9.4, 8.9)
grow_light_2_plant_food_A <- c(8.5, 9.2, 8.9)
grow_light_1_plant_food_B <- c(7.1, 7.2, 8.5)
grow_light_2_plant_food_B <- c(5.5, 5.8, 7.6)

# Create a matrix for plant growth data
task8 <- matrix(c(rep("grow_light_1", 6), rep("grow_light_2", 6), rep("food_A", 3), rep("food_B", 3), rep("food_A", 3), rep("food_B", 3) , grow_light_1_plant_food_A , grow_light_1_plant_food_B, grow_light_2_plant_food_A , grow_light_2_plant_food_B ),ncol=3)

# Convert the matrix to a data frame 
df_8 <- data.frame(task8)

names(df_8) <- c('grow_light', 'food', 'improvements')
```

### Step 1. Stating the hypotheses and identify the claim

The hypothesis for the above will be as follow:

**Null Hypothesis** =

1.  There is no interaction between the two factors (Grow-light and Plant food) in affecting plant growth.

2.  There is no difference in mean growth with respect to light (Grow-light).

3.  There is no difference in mean growth with respect to plant food.

**Alternative Hypothesis** =

1.  There is an interaction between the two factors (Grow-light and Plant food) in affecting plant growth.

2.  There is a difference in mean growth with respect to light (Grow-light).

3.  There is a difference in mean growth with respect to plant food.

### Step 2. Finding the critical value

```{r}
alpha <- 0.05

# Degrees of freedom
dof1 <- 1  # for both main effects (grow light and plant food)
dof2 <- 8  # total observations minus the number of levels

# Critical value for grow light
critical_value_grow_light <- qf(1 - alpha, dof1, dof2)

# Critical value for plant food
critical_value_plant_food <- qf(1 - alpha, dof1, dof2)

critical_value_grow_light
critical_value_plant_food

```

### Step 3. Computing the test results

```{r}
# Perform a two-way ANOVA
anova_result_8 <- aov(improvements ~ grow_light * food, data = df_8) # (OpenAI, 2024)


# Print the ANOVA summary
summary(anova_result_8)
```

### Step 4 - Analyzing the results

**Interpretation:**

-   **grow_light**:

    -   **F value**: 3.68051

    -   Since 3.68051 is less than the critical value of 5.318, the effect of grow light strength on plant growth is not statistically significant at the 0.05 level.

-   **food (Plant Food)**:

    -   **F value**: 24.56230

    -   Since 24.56230 is greater than the critical value of 5.318, the effect of plant food supplement on plant growth is statistically significant at the 0.05 level.

-   **grow_light:food (Interaction between Grow Light and Food)**:

    -   **F value**: 1.43770

    -   Since 1.43770 is less than the critical value of 5.318, the interaction between grow light strength and plant food type does not significantly affect plant growth at the 0.05 level.

### Step 5 - Summary

In summary, based on the critical F value (5.318), only the type of plant food supplement (food) has a significant impact on plant growth. Neither the grow light strength nor the interaction between grow light strength and plant food type shows a significant effect.

The type of plant food supplement significantly affects plant growth, while the grow light strength and the interaction between grow light strength and plant food type do not have a significant effect at the 0.05 level.

------------------------------------------------------------------------

# Task 9

### 9.1 Downloading and importing dataset

```{r}
# Loading necessary libraries
library(ggplot2)
library(dplyr)
library(corrplot)
library(tibble)

# Load the data
data <- read.csv('baseball-3.csv')

```

### 9.2 Exploratory Data Analysis

#### 1. Exploring dataset using descriptive statistics

```{r}

# Summary statistics
summary(data)

# Checking for missing values
cat("No. of NA's =",sum(is.na(data)))


```

-   The above code gives us the ***structure*** of each attribute along with ***measures of central tendency**.*

#### 2. Plots

#### Correlation Matrix

```{r}

# Correlation matrix for numerical variables
numeric_data <- data %>% select(Year, RS, RA, W, OBP, SLG, BA, G, OOBP, OSLG	,RankSeason,	RankPlayoffs)
cor_matrix <- cor(numeric_data, use = "complete.obs") # Excluding missing values
corrplot(cor_matrix, method = 'circle')

# Adding title below the plot
mtext("Plot 1 - Correlation Matrix for Numeric Variables", side = 1, line = 4, cex = 0.8)

```

**Interpretation of Plot 1 - Correlation Matrix for numerical variables**:

The correlation heatmap illustrates the relationships between different pairs of variables. Large, dark blue circles indicate a strong positive correlation, while large, dark red circles indicate a strong negative correlation. For instance, there is a strong positive correlation between Runs Scored (RS) and Wins (W), suggesting that teams that score more runs tend to win more games. On the other hand, there's a notable negative correlation between Runs Allowed (RA) and Wins (W), indicating that allowing fewer runs is associated with more wins.

------------------------------------------------------------------------

#### Distribution of wins

```{r}
# Visualizing the distribution of Wins (W)
ggplot(data, aes(x = W)) + 
  geom_histogram(binwidth = 5, fill = 'blue', color = 'black') +
  labs(title = 'Plot 2 - Distribution of Wins', x = 'Wins', y = 'Frequency')

```

**Interpretation of Plot 2 - Distribution of Wins**

The histogram shows the distribution of wins across the dataset. It resembles a bell-shaped curve, which suggests a normal distribution, with most teams achieving wins in the middle range of the distribution. The highest frequency falls in the 70-90 win range, which is likely close to the mean of the distribution.

------------------------------------------------------------------------

#### Scatter plot for RS vs Wins

```{r}

# Scatter plot for Runs Scored (RS) vs Wins (W)
ggplot(data, aes(x = RS, y = W)) + 
  geom_point() +
  geom_smooth(method = 'lm', col = 'blue') +
  labs(title = 'Plot 3 - Runs Scored vs Wins', x = 'Runs Scored (RS)', y = 'Wins (W)')


```

**Interpretation of Plot 3 - Runs Scored vs Wins**

This scatter plot displays the relationship between Runs Scored (RS) and Wins (W), with a line of best fit indicating a positive linear relationship. This suggests that as the number of runs scored by a team increases, so does the number of wins, which is consistent with the strong positive correlation seen in the heatmap.

------------------------------------------------------------------------

#### Boxplot comparing OBP for teams that made and didn't make the playoffs

```{r}
# Boxplot comparing OBP for teams that made and didn't make the playoffs
ggplot(data, aes(x = as.factor(Playoffs), y = OBP)) + 
  geom_boxplot() +
  labs(title = 'Plot 4 - On-Base Percentage by Playoff Appearance', x = 'Playoffs', y = 'On-Base Percentage (OBP)')

```

**Interpretation of Plot 4 - On-Base Percentage by Playoff Appearance**

The boxplot compares the On-Base Percentage (OBP) between teams that made the playoffs (1) and those that did not (0). Teams that made the playoffs tend to have a higher OBP, as shown by the median line within the box, which is higher for the '1' category. The spread of OBP for playoff teams is also slightly wider, indicating more variability among these teams.

------------------------------------------------------------------------

```{r}
# 6. Rank and Wins
ggplot(data, aes(x = RankSeason, y = W, size = G)) +
  geom_point() +
  labs(title = "Plot 5 - Season Rank vs Wins")
```

**Interpretation of Plot 5 - Season Rank vs Wins**

The scatter plot illustrates the relationship between the season rank (RankSeason) and the number of wins (W) for various teams, with the point size indicating the number of games (G) played. Higher-ranked teams tend to have more wins, but there is considerable overlap, indicating other factors may influence win totals.There are fewer data points for lower ranks, and there is no clear trend showing a relationship between the number of games played and wins or rank.

### Overall

The plots above give an insight into how different attributes are distributed and also how two attributes relate to each other.

------------------------------------------------------------------------

### 9.3 Chi-Square Goodness-of-Fit test

Form the vectors of $expected$ probabilities and $observed$ frequencies, then run the test

```{r}
# Extract decade from year
data$Decade <- data$Year - (data$Year %% 10)
# Create a wins table by summing the wins by decade
wins <- data %>%
group_by(Decade) %>%
summarize(wins = sum(W)) %>%
as.tibble()

number_of_decades <- length(unique(data$Decade))

# Expected proportions for each decade
expected_wins <- rep(1/number_of_decades, number_of_decades)


```

### Step 1 - State the hypotheses and identify the claim

-   **Null hypothesis** **H0**: There is no difference in the number of wins by decade.
-   **Alternative hypothesis** **H1**: There is a difference in the number of wins by decade.

### Step 2 - Calculate critical value at alpha = 0.05

```{r}
critical_value <- qchisq(p = 0.95, df = length(wins$wins) - 1)
cat("Critical value at alpha = 0.05 = ",critical_value)

```

### Step 3 - Compute test value

```{r}
chi_square_test<- chisq.test(x = wins$wins, p = expected_wins)
test_value <- chi_square_test$statistic
chi_square_test
cat("Test value = ", test_value)

```

### Step 4 - Make the decision to reject or not reject the null hypothesis

The decision rule when the $p$-value is less than $\alpha$ we should reject the null hypothesis.

$p$-value \< $\alpha$ i.e. 0.00000000000000022 $<$ 0.05,so we reject $\it{H_{0}}$

The decision rule for test value is that If the ***test value*** exceeds the ***critical value***, you would reject the null hypothesis.

Test value \> critical value i.e 9989.536 \> 11.0705, so we reject $\it{H_{0}}$

### Step 5 - Does comparing the critical and test values provide the same result as comparing the p-value from R with the significance level?

Yes, comparing the ***critical*** and ***test value*** provide the same result as comparing the ***p-value*** with ***significance level***. In both the cases out final decision is to reject the Null hypothesis.

------------------------------------------------------------------------

### 9.4 Download and import the dataset

```{r}
df_9 <- read.csv("crop_data-3.csv")
```

### 9.5 Two-way ANOVA

```{r}

# Convert 'density' and 'fertilizer' to factors
df_9$density <- as.factor(df_9$density)
df_9$fertilizer <- as.factor(df_9$fertilizer)

# Perform Two-way ANOVA
anova_result <- aov(yield ~ density * fertilizer, data = df_9) # (OpenAI,2024)

# Display the ANOVA table
summary(anova_result)

```

### a. Stating the hypothesis

**For density:**

-   **H0**: The means of **yield** are the same across all levels of **density**.

-   **H1**: At least one level of **density** has a mean of **yield** that is different from the others.

**For fertilizer:**

-   **H0**: The means of **yield** are the same across all levels of **fertilizer**.

-   **H1**: At least one level of **fertilizer** has a mean of **yield** that is different from the others.

**For the interaction of density and fertilizer:**

-   **H0**: There is no interaction effect between **density** and **fertilizer** on **yield**.

-   **H1**: There is an interaction effect between **density** and **fertilizer** on **yield**.

### b. Decision

1.  **Density**:

    -   **p-value (Pr(\>F)):** 0.000186

    The ***p-value*** for density is less than 0.05 (significance level). We reject the null hypothesis (H0) that different densities do not affect yield, and accept the alternative hypothesis (H1) that different densities do affect yield.

2.  **Fertilizer**:

    -   **p-value (Pr(\>F)):** 0.000273

    Similarly, the ***p-value*** for fertilizer is also **less** than **0.05**. We reject the null hypothesis that different fertilizers do not affect yield, and accept the alternative hypothesis (H1) that different fertilizers do have an effect on yield.

3.  **Interaction of Density and Fertilizer**:

    -   **p-value (Pr(\>F)):** 0.532500

    The ***p-value*** for the interaction term is ***greater*** than 0.05, which suggests that there is no significant interaction effect between density and fertilizer on yield. We do not reject the null hypothesis for the interaction, indicating that there is not enough evidence to support the claim that the effect of fertilizer on yield depends on the density, or vice versa.

**Conclusion**: Based on the two-way **ANOVA test** conducted above, both density and fertilizer individually have a **significant** effect on yield. However, the combination of density and fertilizer **does not have a significant** interaction effect on yield. This means while each factor separately influences yield, the influence of one factor does not depend on the level of the other factor.

***\* Got an understanding of two-way anova from -*** ("Two-way ANOVA in R," n.d.)

------------------------------------------------------------------------

# Conclusion

The analyses within this module reveal diverse findings. In medical research, the distribution of blood types in a hospital did not significantly differ from the general population. Airline delay reasons varied significantly from government statistics. Movie attendance showed a dependency on ethnicity. There was a significant relationship between rank and branch for women in the military. No notable differences were found in the sodium content among various foods, nor in the sales among different company categories. Educational spending showed no significant regional differences. Lastly, in agriculture, the impact of various growth factors on plants was elucidated. Overall, these insights demonstrate the power of statistical analysis in uncovering meaningful patterns across various sectors.

------------------------------------------------------------------------

# **References**

1.  OpenAI. (2021). ChatGPT (Version 3.5). OpenAI.<https://chat.openai.com/>

2.  Two-way ANOVA in R. (n.d.). Retrieved from <https://statsandr.com/blog/two-way-anova-in-r/>
